---
title: "Workshop 8 TEMPLATE"
subtitle: "9 AM"
author: "Linda Herrera"
date: "2026-02-27"
format:
  html:
    toc: true # use this to display a table of contents
execute:
  message: false # use this to make sure messages don't show up
  warning: false # use this to make sure warnings don't show up
---

[deployed html url](https://herreralinda847.github.io/workshop-08_winter-2026/code/workshop-code.html)

In this workshop, we will answer the question: How do specific leaf area, water treatment, and species influence plant mass?

**Specific leaf area** is a continuous variable measured in cm^2^/g.

**Water treatment** is a categorical variable (i.e. a *factor*) with 2 levels: drought stressed (DS) and well watered (WW).

**Species** is a categorical variable (again, a *factor*) with 7 levels.

| Species name                 | Species code | Common name        |
|------------------------------|--------------|--------------------|
| *Encelia californica*        | ENCCAL       | Bush sunflower     |
| *Eschscholzia californica*   | ESCCAL       | California poppy   |
| *Penstemon centranthifolius* | PENCEN       | Scarlet bugler     |
| *Grindelia camporums*        | GRINCAM      | Gumweed            |
| *Salvia leucophylla*         | SALLEU       | Purple sage        |
| *Stipa pulchra*              | STIPUL       | Purple needlegrass |
| *Lotus scoparius*            | LOTSCO       | Deerweed           |

# 1. Set up

```{r}
#| label: packages and data

# read in your packages here
library(tidyverse) # general use
library(here) # file organization
library(janitor) # cleaning data frames
library(readxl) # reading excel files
library(scales) # modifying axis labels
library(ggeffects) # getting model predictions
library(MuMIn) # model selection

# read in your data here

drought_exp <- read_xlsx(
  path = here(
    "data", 
    "Valliere_etal_EcoApps_Data.xlsx"
    ), 
  sheet = "First Harvest"
)

```

```{r}
#| label: plot colors and aesthetics

# storing colors to use for species
lotsco_col <- "#E69512"
pencen_col <- "#D6264F"
salleu_col <- "#6D397D"
enccal_col <- "#3A5565"
stipul_col <- "#3F564F"
esccal_col <- "#515481"
gricam_col <- "#6C91BD"

# storing colors to use for water treatments
ds_col <- "#A62F03"
ww_col <- "#045CB4"

# storing a ggplot theme (that will be used for all ggplots)
theme_set(theme_bw())
```

# 2. Clean data

Data source: Valliere, Justin; Zhang, Jacqueline; Sharifi, M.; Rundel, Philip (2019). Data from: Can we condition native plants to increase drought tolerance and improve restoration success? \[Dataset\]. Dryad. <https://doi.org/10.5061/dryad.v0861f7>

```{r}
#| label: data cleaning

# creating a new object called 'drought_exp_clean'
drought_exp_clean <- drought_exp |> 
  # cleaning up column names
  clean_names() |> 
  # created new column called 'species_name' the full names of the plants
  mutate(species_name = case_match(
    species,
    "ENCCAL" ~ "Encelia californica", # bush sunflower
    "ESCCAL" ~ "Eschscholzia californica", # California poppy
    "PENCEN" ~ "Penstemon centranthifolius", # Scarlet bugler
    "GRICAM" ~ "Grindelia camporum", # Gumweed
    "SALLEU" ~ "Salvia leucophylla", # purple sage
    "STIPUL" ~ "Stipa pulchra", # purple needlegrass
    "LOTSCO" ~ "Lotus scoparius" # deerweed
  )) |> 
  # adding a column for treatment and changing abbreviations to corresponding with full names
  mutate(water_treatment = case_match(
    water,
    "WW" ~ "Well watered",
    "DS" ~ "Drought stressed"
  )) |> 
  # treating species name and treatments as a factor and seeting factor level order
  mutate(species_name = as_factor(species_name),
         species_name = fct_relevel(species_name,
                                    "Lotus scoparius",
                                    "Penstemon centranthifolius",
                                    "Salvia leucophylla",
                                    "Encelia californica",
                                    "Stipa pulchra",
                                    "Eschscholzia californica",
                                    "Grindelia camporum")) |> 
  mutate(water_treatment = as_factor(water_treatment),
         water_treatment = fct_relevel(water_treatment,
                                       "Drought stressed",
                                       "Well watered")) |> 
  # selecting columns of interest
  select(species_name, water_treatment, sla, total_g)
```

Insert code to double check the structure of the data frame here.

```{r}
#| label: checking structure 

str(drought_exp_clean)

```

Insert code to display 10 random rows from the data frame here.

```{r}
#| label: displaying rows in data frame

slice_sample(
  drought_exp_clean,
  n = 10
)

```

# 3. Visualizing data

## What is the relationship between SLA and plant mass?

Make a scatterplot to visualize SLA (the predictor) on the x-axis, and total mass (the response) on the y-axis.

Label the axes with units where appropriate.

Add a title describing the relationship (e.g. as SLA increases, total mass \_\_\_\_\_\_).

```{r}
#| label: ggplot

ggplot(data = drought_exp_clean,
       aes(x = sla,
           y = total_g)) + 
  # first layer: points to represent individual observations 
  geom_point(color = "tomato") + 
  # labeling the x- and y- axes
  labs(x = "Specific leaf area (cm\U00B2/g)",
       y = "Total mass",
       title = "As SLA increases, total mass increases")

```

## What are the differences in total mass between water treatments?

Make a jitterplot to visualize the differences in plant mass between water treatments.

Label the axes with units where appropriate.

Add a title describing the differences in (mean or median) plant mass between water treatments.

```{r}
#| label: water treatment and mass plot

ggplot(data = drought_exp_clean,
       aes(x = water_treatment,
           y = total_g,
           color = water_treatment)) + 
  geom_jitter(height = 0, 
              width = 0.3) +
  scale_color_manual(values = c(
    "Drought Stressed" = ds_col,
    "Well watered" = ww_col)) + 
  theme(legend.position = "none")
  labs(x = "Water treatment",
       y = "Total mass (g)",
       title = "On average plant mass is well watered treatment is higher than plant mass in drought stressed treatment") + 
  theme_minimal()

```

## What are the differences in total mass between species?

Make a jitterplot to visualize the differences in plant mass between species.

Label the axes with units where appropriate.

Add a title describing the differences in (mean or median) plant mass between species.

Bonus: order the x-axis by decreasing mean plant mass *and* wrap the long species names using the `label_wrap()` function from the `scales` package.

```{r}
#| label: species and plant mass plot 

ggplot(data = drought_exp_clean,
       aes(x = species_name, 
           y = total_g,
           color = species_name)) + 
  geom_jitter(height = 0, 
              width = 0.2) + 
  scale_color_manual(values = c(
     "Lotus scoparius" = lotsco_col,
     "Penstemon centranthifolius" = pencen_col,
     "Salvia leucophylla" = salleu_col,
     "Encelia californica" = enccal_col,
     "Stipa pulchra" = stipul_col,
     "Eschscholzia californica" = esccal_col,
     "Grindelia camporum" = gricam_col
  )) + 
  scale_x_discrete(labels = label_wrap(10)) + 
  labs(x = "Species name",
       y = "Total mass(g)",
       title = "Plant mass differs across species") + 
  theme(legend.position = "none")

```




# 4. Fitting models

8 models total:

| Model number | SLA | Water treatment | Species |       Predictor list        |
|--------------|:---:|:---------------:|:-------:|:---------------------------:|
| 0            |     |                 |         | no predictors (null model)  |
| 1            |  X  |        X        |    X    | all predictors (full model) |
| 2            |  X  |        X        |         |   SLA and water treatment   |
| 3            |  X  |                 |    X    |       SLA and species       |
| 4            |     |        X        |    X    | water treatment and species |
| 5            |  X  |                 |         |             SLA             |
| 6            |     |        X        |         |       water treatment       |
| 7            |     |                 |    X    |           species           |

## Model fitting

```{r}
#| label: model fitting

# model 0: null model


# model 1: all predictors


# model 2: SLA and water treatment


# model 3: SLA and species


# model 4: water treatment and species


# model 5: SLA


# model 6: water treatment


# model 7: species

```

## Model selection

Insert code to use the `AICc()` function to calculate the AIC for each model.

Arrange by decreasing AIC.

## Model diagnostics

Insert code to look at the diagnostic plots for the best model.

## Summary

Insert code to display the summary for the best model.

**What are the reference levels?**

**What does the `(Intercept)` represent?**

**What does `water_treatmentWell watered` represent?**

**What does `species_namePenstemon centranthifolius` represent?**

**What does `species_nameEschscholzia californica` represent?**

## Stop and think: what does this model mean?

**What is the best model?**

\[insert your response here\]

**How much variation in the response (total mass, in grams) does this model explain?**

\[insert your response here\]

**How do we interpret the effects of the predictors on the response variable (again, total mass in grams)?**

\[insert your response here\]

# 4. Model predictions

Insert code to generate model predictions for the best model.

Rename the `x` column and `group` column to match the original data frame (`drought_exp_clean`).

Look at the object before plotting.

# 5. Final figure

Create a clean, publication ready figure depicting predictions for the best model and the underlying data.

# 6. Writing

## Figure caption

## Basic components of test

**predictors**

**response**

**test**

**distribution**

**parameters (degrees of freedom)**

**test statistic**

**R^2^**

**p-value**

**significance level**

**reference level(s)**

## Written communication

### Methods

#### Goal of analysis

#### Model selection

#### Model diagnostics

### Results
